---
title: "PriyasMetabolomics"
format: html
---ee
---

```{r}
library(dplyr)
library(randomForest)
library(rfPermute)
library(vegan)
library(ggplot2)
```

```{r}
oculina <- read.csv("Data/Metabolomics/oculina_superset_final.csv")
ocu_meta <- read.csv("Data/Metabolomics/oculina_metabolomics_metadata.csv")
```

```{r}
print(oculina$chla_pg_sym)
```

PCoA w/ outliers **on oculina_final:**

```{r}

metapheno_fin <- oculina_final$bleaching
meta_id_fin <- oculina_final$id


braycurtis_fin <- vegdist(num_df_final, method = "bray")
pcoa_result_fin <- cmdscale(braycurtis_fin, eig = TRUE)
eig_vals_fin <- pcoa_result_fin$eig
variance_explained_fin <- eig_vals_fin/sum(eig_vals_fin) * 100

pcoa_df_ocu_fin <- data.frame(
  PC1 = pcoa_result_fin$points[, 1],
  PC2 = pcoa_result_fin$points[, 2],
  Phenotype = metapheno_fin,
  ID = meta_id_fin
)


pcoa_oculina_fin <-ggplot(data = pcoa_df_ocu_fin, aes(x = PC1, y = PC2, color = Phenotype)) + 
  geom_point() +
 
  scale_color_manual(values = c("B" = "orange", "NB" = "blue")) +
  
  theme_classic(base_size = 8) +
  theme(plot.margin=ggplot2::margin(t=5,b=2,r=5,l=5,unit="pt"),
        ) +

  labs(
       x = paste("PC1 ", sprintf("%.1f%%", variance_explained_fin[1])),
       y = paste("PC2 ", sprintf("%.1f%%", variance_explained_fin[2])),
       ) +
  theme(plot.margin=ggplot2::margin(t=8,b=8,r=15,l=20,unit="pt"))



   
pcoa_oculina_fin
```

PCoA w/out outliers **on oculina_final:**

```{r}
num_df_fin_no_out <- num_df_final[-c(32, 34, 57, 61),]
metapheno_fin_no_out <- oculina_final[-c(32, 34, 57, 61),]$bleaching
meta_id_fin_no_out <- oculina_final[-c(32, 34, 57, 61),]$id


braycurtis_fin_no_out <- vegdist(num_df_fin_no_out, method = "bray")
pcoa_result_fin_no_out <- cmdscale(braycurtis_fin_no_out, eig = TRUE)
eig_vals_fin_no_out <- pcoa_result_fin_no_out$eig
variance_explained_fin_no_out <- eig_vals_fin_no_out/sum(eig_vals_fin_no_out) * 100

pcoa_df_ocu_fin_no_out <- data.frame(
  PC1 = pcoa_result_fin_no_out$points[, 1],
  PC2 = pcoa_result_fin_no_out$points[, 2],
  Phenotype = metapheno_fin_no_out,
  ID = meta_id_fin_no_out
)


pcoa_oculina_fin_no_out <-ggplot(data = pcoa_df_ocu_fin_no_out, aes(x = PC1, y = PC2, color = Phenotype)) + 
  geom_point() +
 
  scale_color_manual(values = c("B" = "orange", "NB" = "blue")) +
  
  theme_classic(base_size = 8) +
  theme(plot.margin=ggplot2::margin(t=5,b=2,r=5,l=5,unit="pt"),
        ) +

  labs(
       x = paste("PC1 ", sprintf("%.1f%%", variance_explained_fin_no_out[1])),
       y = paste("PC2 ", sprintf("%.1f%%", variance_explained_fin_no_out[2])),
       ) +
  theme(plot.margin=ggplot2::margin(t=8,b=8,r=15,l=20,unit="pt"))



   
pcoa_oculina_fin_no_out
```

PERMANOVA with outliers **on 'oculina_final'**:

```{r}

permanova_fin <- adonis2(braycurtis_fin ~ bleaching, data = oculina_final, permutations = 999)

# Display the results
print(permanova_fin)

```

PERMANOVA without outliers **on 'oculina_final'**:

```{r}
permanova_fin_no_out <- adonis2(braycurtis_fin_no_out ~ bleaching, data = oculina_final[-c(32, 34, 57, 61),], permutations = 999)

# Display the results
print(permanova_fin_no_out)
```

random forest model to predict bleaching phenotype using all metabolites **on 'oculina'**:

```{r}

num_df_phen <- num_df |>
  mutate(Phenotype = oculina$bleaching)


num_df_final <- ocu_final[,-c(1:39, 16397:16426)]
num_df_final_phen <- num_df_final |>
  mutate(Phenotype = ocu_final$bleaching)

typeof(num_df_final_phen$Phenotype)

num_df_final_phen$Phenotype <- as.factor(num_df_final_phen$Phenotype)

class(num_df_final_phen$Phenotype)

#rf_phen <- randomForest(Phenotype ~., data = num_df_final_phen, ntree = 500)

#rf_phen
```

random forest model to predict chlorophyll a using all metabolites **on oculina_final:**:

```{r}

num_df_final_chla <- num_df_final |>
  mutate(chla = ocu_final$ug_chla_cm)



typeof(num_df_final_chla$chla)

class(num_df_final_chla$chla)

rf_chla <- randomForest(chla ~., data = num_df_final_chla, ntree = 500)

rf_chla

```

random forest model to predict symbiont density using all metabolites **on oculina_final:**:

```{r}
num_df_final_sym <- num_df_final |>
  mutate(sym = ocu_final$sym_cm2)



typeof(num_df_final_sym$sym)

class(num_df_final_sym$sym)

rf_sym <- randomForest(sym ~., data = num_df_final_sym, ntree = 500)

rf_sym
```

random forest model to predict \[physio #3\] using all metabolites:

```{r}

```

rfPermute of phenotype **on oculina_final:**

```{r}

rfp_phen <- rfPermute(Phenotype ~., data = num_df_final_phen, na.action = na.omit, ntree = 500, num.rep = 100)

rfp_phen

rfp_table <- as.data.frame(importance(rfp_phen))  # Convert to a data frame
rfp_table <- tibble::rownames_to_column(rfp_table, var = "Feature")
print(rfp_table)  # Display in console

significant_features <- rfp_table %>% filter(B.pval < 0.05 & NB.pval < 0.05)  # Adjust p-value threshold as needed

print(significant_features)

```

log2 fold change of all significant features **on oculina_final:**:

```{r}

#write_csv(significant_features, file = "sig_feat_phenotype_rfpermute.csv")
sig_feat_phen <- read.csv("sig_feat_phenotype_rfpermute.csv")

sig_rows <- sig_feat_phen$Feature

num_df_sig_phen <- data.matrix(num_df_final |>
  select(all_of(sig_rows)))

log2FC <- sapply(sig_rows, function(met){
  
  B_vals  <- (num_df_sig_phen[ocu_final$bleaching == "B", met])
  NB_vals <- (num_df_sig_phen[ocu_final$bleaching == "NB", met])

  mean_B  <- mean(B_vals,  na.rm = TRUE)
  mean_NB <- mean(NB_vals, na.rm = TRUE)

  log2(mean_B / mean_NB)
})

log2FC_df <- data.frame(
  metabolite = sig_feat_phen,
  log2FC = log2FC
)



```

variable importance plot from rfpermute by phenotype **on oculina_final:**:

```{r}
library(ggpattern)

sig_feat_phen_joined <- log2FC_df |>
  inner_join(feats_with_classes, join_by(metabolite.Feature == metabolite)) |>
  select(metabolite.Feature, 
         `metabolite.MeanDecreaseAccuracy`, 
         compound_class, 
         origin, 
         npc_number_superclass)

sig_feat_phen_plot <- sig_feat_phen_joined |>
  arrange(desc(metabolite.MeanDecreaseAccuracy)) |>
  slice_head(n = 50)

ggplot(sig_feat_phen_plot, 
       aes(x = reorder(metabolite.Feature, `metabolite.MeanDecreaseAccuracy`), 
           y = `metabolite.MeanDecreaseAccuracy`,
           fill = compound_class
           )) +
  geom_col(
    ) +
  
  scale_fill_manual(values = c("#C969A1FF", "#CE4441FF", "#EE8577FF", "#EB7926FF", "#FFBB44FF", "#859B6CFF", "#62929AFF", "#004F63FF", "#122451FF")
) +
  
  coord_flip() +  # Flip for better readability
  labs(title = "Variable Importance after rfPermute for Phenotype Classification",
       x = "Feature",
       y = "Mean Decrease in Accuracy") +
  theme_minimal()

```

rfpermute regression by chlorophyll **on oculina_final:**:

```{r}
rfp_chlor <- rfPermute(chla ~., data = num_df_final_chla, na.action = na.omit, ntree = 500, num.rep = 100)

rfp_chlor

rfp_table_chlor <- as.data.frame(importance(rfp_chlor))  # Convert to a data frame
rfp_table_chlor <- tibble::rownames_to_column(rfp_table_chlor, var = "Feature")
print(rfp_table_chlor)  # Display in console

#significant_features_chlor <- rfp_table_chlor %>% filter(B.pval < 0.05 & NB.pval < 0.05)  # Adjust p-value threshold as needed

#print(significant_features_chlor)
```

```{r}
write.csv(rfp_table_chlor, file = "rfp_chlor.csv")
```

```{r}

rfp_chlor <- read.csv("rfp_chlor.csv")
significant_features_chlor <- rfp_chlor %>% filter(`X.IncMSE.pval` < 0.05)
print(significant_features_chlor)
```

log2FC for significant features in chlorophyll; to determine B and NB over representation for these features **on oculina_final:**

```{r}
sig_rows_chlor <- significant_features_chlor$Feature

num_df_sig_chlor <- data.matrix(num_df_final |>
  select(all_of(sig_rows_chlor)))

log2FC_chlor <- sapply(sig_rows_chlor, function(met){
  
  B_vals_chlor  <- (num_df_sig_chlor[ocu_final$bleaching == "B", met])
  NB_vals_chlor <- (num_df_sig_chlor[ocu_final$bleaching == "NB", met])

  mean_B_chlor  <- mean(B_vals_chlor,  na.rm = TRUE)
  mean_NB_chlor <- mean(NB_vals_chlor, na.rm = TRUE)

  log2(mean_B_chlor / mean_NB_chlor)
})

log2FC_df_chlor <- data.frame(
  metabolite = significant_features_chlor,
  log2FC = log2FC_chlor
)

log2FC_df_chlor_upInB <- log2FC_df_chlor |>
  filter(
    log2FC >= 0
  )

write.csv(log2FC_df_chlor_upInB, file = "rfp_chlor_sig_overinB.csv")
```

```{r}
library(ggpattern)


significant_features_chlor_plot <- log2FC_df_chlor |>
  arrange(desc(`metabolite.X.IncMSE`)) |>
  mutate(Sign = if_else(log2FC > 0, "Aposymbiotic", "Symbiotic")) |>
  slice_head(n = 50) 

significant_features_chlor_plot <- significant_features_chlor_plot |>
  inner_join(feats_with_classes, join_by(metabolite.Feature == metabolite)) |>
  select(metabolite.Feature, 
         `metabolite.X.IncMSE`, 
         Sign, 
         compound_class, 
         origin, 
         npc_number_superclass)

significant_features_chlor_plot <- significant_features_chlor_plot |>
  mutate(Symbol  = case_when(origin == "host & symbiont" ~ "\U2605",
                             origin == "symbiont" ~ "\U2713",
                             origin == "symbiosis-specific" ~ "\U2660"))


```

```{r, fig.height = 10, fig.width = 10}
ggplot(significant_features_chlor_plot, 
       aes(x = reorder(metabolite.Feature, `metabolite.X.IncMSE`), 
           y = `metabolite.X.IncMSE`,
           fill = compound_class,
           pattern = Sign
           )) +
 # scale_fill_manual(values = c("Bleached" = "orange", "Non-Bleached" = "blue")) +
  geom_col_pattern(
  pattern_density = 0.2,
  pattern_spacing = 0.03
  ) +
  
  
  scale_fill_manual(values = c("#FBE183FF", "#F4C40FFF", "#FE9B00FF", "#D8443CFF", "#9B3441FF", "#DE597CFF", "#E87B89FF", "#E6A2A6FF", "#AA7AA1FF", "#9F5691FF", "#633372FF", "#1F6E9CFF", "#2B9B81FF", "#92C051FF")

) +
  geom_text(aes(label = Symbol), vjust = 0, hjust = 0, size = 4) +
  
  coord_flip() +  # Flip for better readability
  labs(title = "Variable Importance after rfPermute for Chloropyll Regression",
       x = "Feature",
       y = "Percent Increase in Mean Squared Error") +
  theme_minimal()
```
