---
title: "Oculina_Metabolomics"
format: html
editor: visual
---

## Setup

### Load libraries

```{r}
#| echo: false
library(tidyverse)
library(here)
library(dplyr)
library(ggplot2)
library(ggpubr)
```

### Read in data

Here is my metadata

```{r}
oc_meta <- read_csv(here("Data", "Metabolomics","oculina_metabolomics_metadata.csv"))
```

This is the data that Ty initially shared with me in July 2025

```{r}
oculina_full <- read_csv(here("Data", "Metabolomics","Oculina_full.csv"))
```

I noticed these issues that did not match my metadata:

The dataset is missing A9-S, did the sample fail or not transfer over?

The dataset is missing SX7-A, did the sample fail or not transfer over?

AX2 and A12 samples are duplicates? I have AX2 bag labeled as A12 in my dataset so I believe this should only be one sample.

There is a duplicate of SX2-A in the dataset and there should only be one. They are labeled as: CuracaoSet2_plate4_SX2-A_20250221104845 and CuracaoSet2_plate4_SX2-A, might just have copied over twice. Although, looking at the actual data, there are different numbers for the different metabolites so I'm not sure.

S16-A and S16S labels are switched (apo vs sym designations and associated metadata are swapped) - which part of the data is actually switched, just the names or the other data? \*\***symbiosis column is swapped, data is correct - MANUALLY EDITING THIS\*\***

There are two S18-A, with the sample names of 03072025_Curacoa_Plate5_S18A and CuracaoSet2_plate4_S18-A. I believe one is supposed to be S18S, do you know which one is which? \***I believe that 03072025_Curacoa_Plate5_S18A is the symbiotic one by sorting several datasets that split evenly apo/sym - Manually editing this\***

Replicate samples that will be helpful checks - just noting this here, no issues with these samples:

SX1 and S1, SX2 and S15, and then AX4 and A15 are duplicate samples from the same coral and I have them both listed in my dataset - will be good to check using these replicates

SO NOW: Looking through data that was shared with me Sept 3rd (to this email: maya.elizabeth.powell\@gmail.com) - BIG raw data

```{r}
superset_raw <- read_csv(here("Data", "Metabolomics","Ty_superset_data_raw.csv"))
```

### Manual edits

```{r}
#manually editing S16-S and -A to have the correct "symbiosis" column values
 superset_raw[superset_raw$sample_id=="S16-S", "symbiosis"] <- "symbiotic"
 superset_raw[superset_raw$sample_id=="S16-A", "symbiosis"] <- "aposymbiotic"
```

```{r}
#manually editing 03072025_Curacoa_Plate5_S18A to be S18-S
superset_raw[superset_raw$sample=="03072025_Curacoa_Plate5_S18A", "sample_id"] <- "S18-S"
superset_raw[superset_raw$sample=="03072025_Curacoa_Plate5_S18A", "symbiosis"] <- "symbiotic"
```

Merging dataframes based on sample_id to see what's missing

```{r}
oculina_raw <- oc_meta %>% left_join(superset_raw, by = "sample_id") #84 samples
```

Samples that we don't have data for are:

SX7-A - **I believe this is accidentally labeled as SX2-A - but we sadly can't tell**

Both SX2 and SX7 samples were run on plate 4 and don't have any different metadata, so I cannot tell them apart - I think this means that both SX2-A samples as well as SX7-A need to be removed from the dataset if we want to do colony tracking. But if we just want to separate them as two unidentified apo samples and not ID/compare them by colony, we can just rename since they are both apo

A9-S - the only thing I can think about this is that there's an A9-5 Hawaiʻi sample that could be it, but it looks like it's from a totally different plate (QE_0924_Lipid_plate19_KARP_T1_A9-5), so removing for now.

### Remove data as needed and clean up dataframe

```{r}
id_remove <- c("A9-S", "SX7-A", "SX2-A")
oculina_clean <- oculina_raw %>%
  filter(!sample_id %in% id_remove)

#can do this to remove x in front of column names - not doing for now just in case but will likely do in the future
#colnames(oculina_clean)<-gsub("x","",colnames(oculina_clean))
```

### Save dataframe

```{r}
write.csv(oculina_clean, here("Data", "Metabolomics", "oculina_clean.csv"), row.names = FALSE)
```
