---
title: "Oculina Metabolomics Initial Data Processing and Physiology"
author: Maya Powell
format: html
editor: visual
---

## Setup

### Load libraries

```{r}
# install packages as needed
library(tidyverse)
library(readr)
library(here)
library(ggplot2)
library(ggpubr)
library(car)
library(rstatix)
library(emmeans)
library(powerjoin)
```

### Read in data

Here is my metadata

```{r}
oc_meta <- read.csv(here("Data", "Metabolomics","oculina_metabolomics_metadata.csv"))
```

This is the data that Ty initially shared with me in July 2025

```{r}
oculina_full <- read.csv(here("Data", "Metabolomics","Oculina_full.csv"))
```

I noticed these issues that did not match my metadata:

The dataset is missing A9-S, did the sample fail or not transfer over?

The dataset is missing SX7-A, did the sample fail or not transfer over?

AX2 and A12 samples are duplicates? I have AX2 bag labeled as A12 in my dataset so I believe this should only be one sample.

There is a duplicate of SX2-A in the dataset and there should only be one. They are labeled as: CuracaoSet2_plate4_SX2-A_20250221104845 and CuracaoSet2_plate4_SX2-A, might just have copied over twice. Although, looking at the actual data, there are different numbers for the different metabolites so I'm not sure.

S16-A and S16S labels are switched (apo vs sym designations and associated metadata are swapped) - which part of the data is actually switched, just the names or the other data? \*\***symbiosis column is swapped, data is correct - MANUALLY EDITING THIS\*\***

There are two S18-A, with the sample names of 03072025_Curacoa_Plate5_S18A and CuracaoSet2_plate4_S18-A. I believe one is supposed to be S18S, do you know which one is which? \***I believe that 03072025_Curacoa_Plate5_S18A is the symbiotic one by sorting several datasets that split evenly apo/sym - Manually editing this\***

Replicate samples that will be helpful checks - just noting this here, no issues with these samples:

SX1 and S1, SX2 and S15, and then AX4 and A15 are duplicate samples from the same coral and I have them both listed in my dataset - will be good to check using these replicates

SO NOW: Looking through data that was shared with me Sept 3rd (to this email: maya.elizabeth.powell\@gmail.com) - BIG raw data

This data has annotations as well for compounds

```{r}
superset_raw <- read.csv(here("Data", "Metabolomics","Ty_superset_data_raw.csv"))
```

### Manual edits

```{r}
#manually editing S16-S and -A to have the correct "symbiosis" column values
 oculina_full[oculina_full$sample_id=="S16-S", "symbiosis"] <- "symbiotic"
 oculina_full[oculina_full$sample_id=="S16-A", "symbiosis"] <- "aposymbiotic"
 superset_raw[superset_raw$sample_id=="S16-S", "symbiosis"] <- "symbiotic"
 superset_raw[superset_raw$sample_id=="S16-A", "symbiosis"] <- "aposymbiotic"
```

```{r}
#manually editing 03072025_Curacoa_Plate5_S18A to be S18-S
oculina_full[oculina_full$sample=="03072025_Curacoa_Plate5_S18A", "sample_id"] <- "S18-S"
oculina_full[oculina_full$sample=="03072025_Curacoa_Plate5_S18A", "symbiosis"] <- "symbiotic"
superset_raw[superset_raw$sample=="03072025_Curacoa_Plate5_S18A", "sample_id"] <- "S18-S"
superset_raw[superset_raw$sample=="03072025_Curacoa_Plate5_S18A", "symbiosis"] <- "symbiotic"
```

Merging dataframes based on sample_id to see what's missing

```{r}
oculina_superset_raw <- oc_meta %>% power_left_join(superset_raw, by = "sample_id", conflict = coalesce_xy) #84 samples
oculina_full_raw <- oc_meta %>% power_left_join(oculina_full, by = "sample_id", conflict = coalesce_xy) #84 samples
```

Samples that we don't have data for are:

SX7-A - **I believe this is accidentally labeled as SX2-A - but we sadly can't tell**

Both SX2 and SX7 samples were run on plate 4 and don't have any different metadata, so I cannot tell them apart - this means that both SX2-A samples as well as SX7-A need to be removed from the dataset if we want to do colony tracking. But if we just want to separate them as two unidentified apo samples and not ID/compare them by colony, we can just rename since they are both apo

A9-S - the only thing I can think about this is that there's an A9-5 Hawaiʻi sample that could be it, but it looks like it's from a totally different plate (QE_0924_Lipid_plate19_KARP_T1_A9-5), so removing for now.

### Remove data as needed and clean up dataframe

```{r}

#add data for general and species columns
oculina_superset_clean <- oculina_superset_raw %>%
  mutate(location = "Radio Island") %>%
  mutate(host_phylum = "Cnidaria") %>%
  mutate(host_class = "Hexacorallia") %>%
  mutate(host_order = "Scleractinia") %>%
  mutate(host_family = "Oculinidae") %>%
  mutate(host_genus = "Oculina") %>%
  mutate(host_species = "arbuscula") %>%
  mutate(bleaching = case_when(frag_sym == "A" ~ 'B',
                               frag_sym == "S" ~ 'NB',))

oculina_clean <- oculina_full_raw %>%
  mutate(location = "Radio Island") %>%
  mutate(host_phylum = "Cnidaria") %>%
  mutate(host_class = "Hexacorallia") %>%
  mutate(host_order = "Scleractinia") %>%
  mutate(host_family = "Oculinidae") %>%
  mutate(host_genus = "Oculina") %>%
  mutate(host_species = "arbuscula") %>%
  mutate(bleaching = case_when(frag_sym == "A" ~ 'B',
                               frag_sym == "S" ~ 'NB',))

#can do this code below to remove x in front of column names
#colnames(oculina_clean)<-gsub("x","",colnames(oculina_clean))
```

### Merge with physiology data

```{r}
physio <- read_csv(here("Data", "Nov_2024", "oculina_nov24_physio.csv"))

#duplicates of metabolomics to duplicate physiology for:
#SX1 = S1 duplicate, SX2 = S15 duplicate, and then AX4 = A15 duplicate

#generate IDs for new data for duplicates
id_map <- tribble(
  ~old_id, ~new_id,
  "S1-A",  "SX1-A",
  "S1-S",  "SX1-S",
  "S15-A", "SX2-A",
  "S15-S", "SX2-S",
  "A15-A", "AX4-A",
  "A15-S", "AX4-S"
)

#create duplicate dataframe to add the phys data to these IDs
physio_duplicates <- physio %>%
  inner_join(id_map, by = c("sample_id" = "old_id")) %>%
  mutate(sample_id = new_id) %>% #rename to sample id to merge
  select(-new_id) #remove new id variable

#add rows into physio dataframe to join with metabolomics
physio_metab <- bind_rows(physio, physio_duplicates)

#create dataframes with physio as well
oculina_superset_final <- oculina_superset_clean %>% power_right_join(physio_metab, by = "sample_id", conflict = coalesce_xy) #79 samples

oculina_final <- oculina_clean %>% power_right_join(physio_metab, by = "sample_id", conflict = coalesce_xy) #79 samples
```

### Remove data based on notes above

```{r}
id_remove <- c("A9-S", "SX7-A", "SX2-A") #removes 4 samples bc SX2-A duplicate

oculina_superset_final <- oculina_superset_final %>% filter(!sample_id %in% id_remove)
oculina_final <- oculina_final %>% filter(!sample_id %in% id_remove)

#76 samples in each which is 80-4 so that makes sense! love it
```

### Save dataframes

@Priya here are the final dataframes for you to use! These have: physiology data, metadata, and metabolomics data. All metadata and physiology data is on the left side so it's easy to see, and physiology data has been duplicated for the duplicate metabolomics samples.

There are two different datasets here - the oculina_superset_final dataframe has 31360 variables (and comes from Ty_superset_raw) while the oculina_final dataframe has 16426 variables (and comes from Oculina_full).

```{r}
#dataframe that has ALL original data - from "Ty superset raw" data
write.csv(oculina_superset_final, here("Data", "Metabolomics", "oculina_superset_final.csv"), row.names = FALSE)

#dataframe that was slightly cleaned up - from "Oculina full"
write.csv(oculina_final, here("Data", "Metabolomics", "oculina_final.csv"), row.names = FALSE)
```

## Physiology

### Read in data

```{r}
sym_chla <- read_csv(here("Data", "Nov_2024", "oculina_nov24_physio.csv"))
```

info about variables - add notes that are helpful to you!

sa_colony = aposymbiotic/symbiotic original designation of the colony

sa_frag = aposymbiotic/symbiotic designation of the FRAGMENT

ug_chla_cm = micrograms of chlorophyll a per cm2 of surface area

sym_cm2 = symbiont density for cm2 of surface area

chla_pg_sym = picograms of chlorophyll a per symbiont

depth = deep (30ft), shallow (15ft)

### Chlorophyll a

Graph

```{r}
#chlorophyll a
chla_nov_depth_sa_box <- ggplot(sym_chla, aes(x = depth_sa, y=ug_chla_cm, fill = depth_sa))+
  geom_boxplot()+
  geom_jitter(alpha=0.8, width=0.2)+
  theme_classic(base_size = 22)+
  labs(x = "Fragment symbiotic state", y = "Chlorophyll a (ug per cm2)")+
  stat_summary(data = sym_chla, aes(x = depth_sa, y = ug_chla_cm), geom = "text", fun = max, vjust = -0.5, size = 8,
               label = c("a", "a", "b"))+ 
  #EDIT LETTERS FOR EACH GRAPH - MAKE SURE THEY MATCH WITH STATS!!
  #SAME LETTERS = SAME (not sig dif from eachother)
  #DIF LETTERS = SIGNFICANTLY DIFFERENT FROM EACHOTHER
  theme(legend.position = "none")+
  ylim(-0.2,7)+ #ELLA - EDIT Y LIMIT IF IT CUTS OFF LETTERS ON THE TOP
  #MAKE SURE YOU DON'T GET THE WARNING THAT IT REMOVES ANY DATA
  scale_fill_manual(values = c("white","lightyellow","brown")) #ella change the colors here as you'd like
chla_nov_depth_sa_box

ggsave(here("Output", "Nov_2024","chla_nov_depth_sa_box.pdf"), device = "pdf", h = 6, w = 10, chla_nov_depth_sa_box)
```

Stats

```{r}
#anova
chla_frag_aov <- aov(ug_chla_cm ~ depth_sa, data = sym_chla)
summary(chla_frag_aov)
Anova(chla_frag_aov)

#pairwise
chla_emm <- emmeans::emmeans(chla_frag_aov, ~depth_sa)
pairs(chla_emm)
```

### Symbiont density

Graph

```{r}
#symbiont density
sym_nov_depth_sa_box <- ggplot(sym_chla, aes(x = depth_sa, y=sym_cm2, fill = depth_sa))+
  geom_boxplot()+
  geom_jitter(alpha=0.8, width=0.2)+
  theme_classic(base_size = 22)+
  labs(x = "Fragment symbiotic state", y = "Symbiont density (cells per cm2)")+
  stat_summary(data = sym_chla, aes(x = depth_sa, y = ug_chla_cm), geom = "text", fun = max, vjust = -0.5, size = 8,
               label = c("a", "b", "b"))+ 
  theme(legend.position = "none")+
  scale_fill_manual(values = c("white","lightyellow","brown")) #ella change the colors here as you'd like
sym_nov_depth_sa_box

ggsave(here("Output", "Nov_2024","sym_nov_depth_sa_box.pdf"), device = "pdf", h = 6, w = 10, sym_nov_depth_sa_box)
```

Stats

```{r}
#anova
sym_frag_aov <- aov(sym_cm2 ~ depth_sa, data = sym_chla)
summary(sym_frag_aov)
Anova(sym_frag_aov)

#pairwise
sym_emm <- emmeans::emmeans(sym_frag_aov, ~depth_sa)
pairs(sym_emm)
```

### Chlorphyll a per symbiont

Graph

```{r}
chla_sym_nov_depth_sa_box <- ggplot(sym_chla, aes(x = depth_sa, y=chla_pg_sym, fill = depth_sa))+
  geom_boxplot()+
  geom_jitter(alpha=0.8, width=0.2)+
  theme_classic(base_size = 22)+
  stat_summary(data = sym_chla, aes(x = depth_sa, y = ug_chla_cm), geom = "text", fun = max, vjust = -0.5, size = 8,
               label = c("a", "a", "b"))+ 
  labs(x = "Fragment symbiotic state", y = "Chlorophyll a (pg per symbiont cell)")+
  theme(legend.position = "none")+
  scale_fill_manual(values = c("white","lightyellow","brown"))
chla_sym_nov_depth_sa_box

ggsave(here("Output", "Nov_2024","chla_sym_nov_depth_sa_box.pdf"), device = "pdf", h = 6, w = 10, chla_sym_nov_depth_sa_box)
```

Stats

```{r}
#anova
chla_sym_frag_aov <- aov(chla_pg_sym ~ depth_sa, data = sym_chla)
summary(chla_sym_frag_aov)
Anova(chla_sym_frag_aov)

#pairwise
chla_sym_emm <- emmeans::emmeans(chla_sym_frag_aov, ~depth_sa)
pairs(chla_sym_emm)
```

### Correlation sym density vs chla

```{r}
#linear plot comparing symbiont density vs chlorophyll a across all 3 groups
sym_chla_scatter <- ggplot(sym_chla, aes(x = sym_cm2, y = ug_chla_cm, fill = depth_sa)) +
  geom_point(color='black', shape = 21) +
  scale_fill_manual(values = c("white","lightyellow","brown"))+
  theme_bw(base_size = 22) +
  geom_smooth(aes(x = sym_cm2, y = ug_chla_cm, group = 1),
              method = "lm", se = TRUE, color = "black", linewidth = 1.1) +
  labs(x = "Symbiont density (cells per cm2)", 
       y = "Chlorophyll a (ug per cm2)", 
       fill = "Fragment symbiotic state")
sym_chla_scatter
ggsave(here("Output", "Nov_2024","sym_chla_scatter.pdf"), device = "pdf", h = 6, w = 10, sym_chla_scatter)

##this indicates that we likely need to remove outliers from this data!! more to do!!
```
